---
title: "Sampling"
author: Jakub Nowosad
date: "`r Sys.Date()`"
---

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

Let's start by attaching the necessary packages and setting the seed.

```{r setup}
library(simsam)
library(terra)
set.seed(2024-11-20)
```

Next, we define the raster grid to sample points/locations from.

```{r}
rast_grid = terra::rast(xmin = 0, xmax = 200, ymin = 0, ymax = 200,
                        ncols = 200, nrows = 200)
```

Alternatively, we can read a raster from a file.

## Sample

The **simsam** package supports several types of sampling, including jittered, clustered, and random with the `sam_field()` function.
All these types of sampling require the `rast_grid` object, the number of samples to create (`size`), and the type of sampling (`type`).

The jittered sampling is a regular sample moved by an amount of noise. 
The `value` parameter controls the amount of jitter.
Importantly, this sampling type ensures that all the samples fall within the sampling window.

```{r}
sample_j1 = sam_field(rast_grid, size = 100, type = "jittered", value = 5)
plot(sample_j1)
sample_j2 = sam_field(rast_grid, size = 100, type = "jittered", value = 20)
plot(sample_j2)
```

The clustered sampling creates samples forming groups of points with similar locations.
The outcome is controlled by the `nclusters` and `value` parameters, where the former controls the number of clusters and the latter the amount of noise (location variance) within each cluster.

```{r}
sample_c1 = sam_field(rast_grid, 100, "clustered", nclusters = 5, value = 5)
plot(sample_c1)
sample_c2 = sam_field(rast_grid, 100, "clustered", nclusters = 5, value = 10)
plot(sample_c2)
```

Random sampling creates samples with no spatial structure.

```{r}
sample_r = sam_field(rast_grid, 100, "random")
plot(sample_r)
```

Moreover, the `sam_field()` function can be used with additional arguments passed to the `terra::spatSample()` function, when the `type` is not specified.
In such cases, the `method` argument can be used to control the sampling method, e.g., `method = "regular"` or `method = "random"` and the `exhaustive` argument to sample only from not NA areas in the input raster.

```{r}
sample_u1 = sam_field(rast_grid, 100, method = "regular")
plot(sample_u1)
sample_u2 = sam_field(rast_grid, 100, method = "random", exhaustive = TRUE)
plot(sample_u2)
```

